---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ShennongData

<!-- badges: start -->
[![R-CMD-check](https://github.com/zerostwo/shennong-data/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/zerostwo/shennong-data/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/zerostwo/shennong-data/branch/main/graph/badge.svg)](https://app.codecov.io/gh/zerostwo/shennong-data)
<!-- badges: end -->

`ShennongData` is a lazy R client for ShennongDB. It discovers resource
metadata first and only downloads data when you explicitly materialize a
query.

## Installation

Install the development version from GitHub:

```{r installation, eval = FALSE}
install.packages("remotes")
remotes::install_github("zerostwo/shennong-data")
```

## Basic usage

Connect to a ShennongDB server and create a metadata-only resource handle:

```{r connect, eval = FALSE}
library(ShennongData)

con <- sn_connect(
  "https://your-shennongdb.example",
  token = Sys.getenv("SHENNONG_TOKEN")
)

x <- sn_load_data("toil", connection = con)
x
sn_schema(x)
sn_layers(x)
```

Choose a view and build a lazy query before collecting data:

```{r query, eval = FALSE}
observations <- sn_obs(x)
expression <- sn_assay(x)

sn_show_query(observations)
result <- collect(observations)
```

Use bounded fetches or artifacts for larger data transfers:

```{r fetch, eval = FALSE}
sn_fetch_data(
  x,
  features = c("YTHDF2", "FTO"),
  layer = "log2_tpm_plus_0.001",
  limit = 100
)
sn_artifacts(x)
```

## Agent and MCP integration

Check the live API contract and discover visible Resources from R:

```{r agent, eval = FALSE}
sn_api_compatibility(con)
sn_resources(con, search = "bulk")
```

The package also includes a read-only stdio MCP server:

```sh
SHENNONG_URL=http://127.0.0.1:18080 \
  Rscript -e 'ShennongData::sn_mcp_serve()'
```

It exposes bounded tools for compatibility checks, Resource discovery and
inspection, identifier resolution, query planning, and small data fetches. See
[`docs/agent-integrations.md`](docs/agent-integrations.md) for client
configuration and safety limits.

See the [pkgdown site](https://zerostwo.github.io/shennong-data/) for the
complete function reference.
