---
title: "Data materialization contracts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data materialization contracts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

ShennongData is a user-facing, read-only client for a Shennong OS or gateway.
It uses a normal session bearer token and never an administrative key. The
package discovers a ShennongDB Resource first, then explicitly materializes a
bounded query or Artifact.

```{r connection}
con <- sn_connect(
  Sys.getenv("SHENNONG_URL", "https://your-shennong-gateway.example"),
  token = Sys.getenv("SHENNONG_TOKEN"),
  project_id = Sys.getenv(
    "SHENNONG_PROJECT_ID",
    "550e8400-e29b-41d4-a716-446655440000"
  )
)
```

With a Project scope, authenticated same-origin requests carry
`X-Shennong-Project-Id`, and query bodies include `project_id` as a gateway
compatibility bridge. A direct DB/public connection remains compatible when
`project_id = NULL`. A governed `project_id` must be a canonical UUID; aliases
and slugs are rejected before network access.

## DataBundle provenance

Every materialized result records the target contract
`shennong.dev/data-bundle/v1`. This does not mean every current ShennongDB
Resource already emits a complete DataBundle:

- `status = "resource_declared"` means Resource metadata declared
  `schema_version = "shennong.dev/data-bundle/v1"`;
- `status = "client_projection"` means ShennongData projected the current
  Resource/query response into the target contract;
- `complete`, `axis_complete`, `missing_features`, and
  `incomplete_reasons` describe whether the projection is safe to consume as a
  complete matrix.

The analysis package uses a different output contract,
`shennong.dev/analysis-result-bundle/v1`. ShennongData produces input data and
provenance; a public Shennong analysis function produces and validates the
analysis result.

```{r provenance}
result <- sn_fetch_data(
  x,
  features = c("YTHDF2", "FTO"),
  layer = "log2_tpm_plus_0.001",
  shape = "long",
  limit = 100
)

sn_result_schema(result)
sn_provenance(result)$data_bundle
sn_is_partial(result)
```

## Shape and zero semantics

The four shapes have explicit contracts:

| Shape | Orientation | Missing coordinate |
| --- | --- | --- |
| `long` | one feature-observation coordinate per row | remains absent |
| `wide` | observations by selected values | remains missing |
| `matrix` | features by observations | zero only when `implicit_zero = TRUE`; otherwise `NA` |
| `sparse` | features by observations (`dgCMatrix`) | structural zero only when `implicit_zero = TRUE` |

For matrix and sparse requests, the client fetches the complete observation
axis when the server advertises the axis endpoint. Without a complete axis,
all-zero observations cannot be recovered, so the result is marked partial.
Batch `missing_features` metadata is also retained as incomplete provenance.
Explicit zero entries are removed from a sparse matrix with `Matrix::drop0()`.

```{r matrices}
dense <- sn_fetch_data(
  x,
  features = c("gene-a", "gene-b"),
  layer = "counts",
  shape = "matrix"
)

sparse <- sn_fetch_data(
  x,
  features = c("gene-a", "gene-b"),
  layer = "counts",
  shape = "sparse"
)
```

## Analysis containers

`sn_as()` supports explicit `matrix` and `sparse` targets and preserves the
same provenance in SummarizedExperiment metadata or Seurat `misc`. Analysis
containers reject partial inputs by default. `allow_partial = TRUE` is an
escape hatch for a user who has reviewed the incomplete reasons; it is not a
completeness assertion.

```{r containers}
se <- sn_as(
  x,
  "SummarizedExperiment",
  features = c("YTHDF2", "FTO"),
  layer = "log2_tpm_plus_0.001"
)

S4Vectors::metadata(se)$shennong$data_bundle
```

The current TOIL layer is exactly `log2_tpm_plus_0.001`. It is transformed
expression, not raw counts, so it is suitable for bulk QC but not for
DESeq2/edgeR count workflows:

```{r shennong-consumer}
bulk_qc <- Shennong::sn_run_bulk(se, workflow = "qc")
Shennong::sn_validate_result(bulk_qc, error = FALSE)$valid
```

## Artifact trust boundary

Same-origin Artifact endpoints may receive the connection bearer token. A
foreign presigned URL never receives it, must use HTTPS, and is followed only
through bounded, manually reconstructed redirects. Automatic credential
forwarding is disabled.

Local paths and `file://` URIs are ignored by default. Trusted-local access
requires both `trusted_local = TRUE` and a `local_root`; normalized paths and
symlink targets must remain under that root.

```{r trusted-local}
sn_download_artifact(
  x,
  artifact = "local-expression",
  path = tempfile(fileext = ".tsv"),
  trusted_local = TRUE,
  local_root = "/srv/shennong/approved-artifacts"
)
```
